Imperative Programming with Variational Effects

### 2018-08-10 paper/paper.tex
mostly done with section 5

\section{Exceptions}

@@

@@

When describing the issues involved in the interaction between mutable state and variation in Chapter \ref{ch:mot}, we noted the need for (1) maintaining a variational context
during execution and (2) using this variational context when interacting with the variable store. This is a solution first discovered by \cite{varwhile} in their work on a
variational model of execution for the \textsc{While} language.

@@

We now provide a sample implementation of a variational store based on the variational map data structure \cite{varwhile, walkingshaw2014variational}.
Variational maps are an efficient map data structure for storing variational values and are easily built using an existing non-variational map. The basic idea
is that for a non-variational map of with values of type \prog{v}, a variational map uses the same underlying data structure but with values of type \prog{V (Maybe v)}.
The variational values in the map must be of type \prog{Maybe v} in order to account for cases where a value is assigned in one variant but not in others.

@@

vmerge :: VOpt v -> Cond -> VOpt v -> -> VOpt v
vmerge (One v) vctx v' = Chc vctx v' (One v)
vmerge v@(Chc c l r) vctx v'@(One _)
  | c => vctx = Chc vctx (One v) (vmerge )
  | vctx => c = Chc c (vmerge l vctx (One v)) r
  | vctx => Not c = Chc c l (vmerge r vctx (One v))
  | otherwise = Chc c (vmerge l vctx (One v)) (vmerge r vctx (One v))
vmerge

vupdate :: Cond -> Var -> V v -> VStore v -> VStore v
vupdate c k v vst = 
  where
    mkOpt = fmap Just
    current = lookup k vst
    case current of
      Nothing -> insert k (Chc c (mkOpt v) (One Nothing)) vst
      Just v -> insert k (vmerge c v current) vst
\end{lstlisting}
\caption{Variational map operations}
\label{fig:varops}
\end{figure}

\begin{figure}[H]
\begin{syntax}
\text{(Store)}\\
S &::=& S(C,x) & \textit{Lookup x in S with context C} \\
\text{(Variational integers)}\\
u &::=& n & \textit{Integer literal} \\
& | & \tagtree{C}{u}{u} & \textit{Choice} \\
\text{(Variational booleans)}\\
v &::=& b & \textit{Boolean literal} \\
& | & \tagtree{C}{v}{v} & \textit{Choice}
\end{syntax}
\end{figure}

\begin{figure}[H]
\ContinuedFloat
\begin{mathpar}
\BigVNum \and
\BigVRef \and
\BigVAdd \and
\BigVAChcOne \and
\BigVAChcTwo \and
\BigVAChcThree \and
\BigVB \and
\BigVNot \and
\BigVAnd \and
\BigVLess \and
\BigVBChcOne \and
\BigVBChcTwo \and
\BigVBChcThree
\end{mathpar}
\label{fig:vimpexpr}
\caption{Big-step semantics of Variational IMP expressions}
\end{figure}

\newcommand{\GetCtx}[2]{\mathit{ctx}(#1,#2)}
\def \gc {\GetCtx{C}{E}}
\def \T {\mathit{true}}
\def \F {\mathit{false}}
\def \BigVUnsat {\infer [V-Unsat] {\mathit{unsat}(\gc)} {(C,S,E,s) \Downarrow_{V} (S,E)}}
\def \BigVSat {\infer [V-Sat] {\mathit{sat}(\gc) \\ (C,S,E,s) \Downarrow_{VS} (S',E')} {(C,S,E,s) \Downarrow_{V} (S',E')}}
\def \BigVSkip {\infer[VS-Skip] { } {(C,S,E,\CCkeyw{skip}) \Downarrow_{VS} (S,E)}}
\def \BigVAssn {\infer[VS-Assn] {(\gc,S,a) \Downarrow_{VA} u} {(C,S,E,x \coloneqq a) \Downarrow_{VS} (S[(\gc,x) \rightarrow u],E)}}
\def \BigVSeq {\infer [VS-Seq] {(C,S,E,s) \Downarrow_{V} (S',E') \\ (C',S',E',s') \Downarrow_{V} (S'',E'')} {(C,S,E,s\ ;\ s') \Downarrow_{VS} (S'',E'')}}
\def \BigVIf {\infer [VS-If] {(\gc,S,e) \Downarrow_{VB} v \\ (C \wedge \mathit{whenTrue}(v),S,E,s) \Downarrow_{V} (S',E') \\ (C \wedge \mathit{whenFalse}(v),S',E') \Downarrow_{V} (S'',E'')} {(C,S,E,\CCkeyw{if}\ e\ \CCkeyw{then}\ s\ \CCkeyw{else}\ s') \Downarrow_{VS} (S'',E'')}}
\def \BigVWhile {\infer [VS-While] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ (C \wedge \mathit{whenTrue}(v),S,E,s) \Downarrow_{V} (S',E') \\ (C \wedge \mathit{whenTrue}(v),S',E',\CCkeyw{while}\ e\ \CCkeyw{do}\ s) \Downarrow_{V} (S'',E'')} {(C,S,E,\CCkeyw{while}\ e\ \CCkeyw{do}\ s) \Downarrow_{VS} (S'',E'')}}
\def \BigVThrow {\infer [VS-Throw] {(\GetCtx{C}{l:E},S,a) \Downarrow_{VA} u} {(C,S,l:E,\CCkeyw{throw}\ a) \Downarrow_{VS} (S,((C,u):l):E)}}
\def \BigVTry {\infer [VS-Try] {(C,S,[]:E,s) \Downarrow_{V} (S',[]:E)} {(C,S,E,\CCkeyw{try}\ s\ \CCkeyw{catch}\ x\ \CCkeyw{with}\ s') \Downarrow_{VS} (S',E)}}
\def \BigVCatch {\infer [VS-Catch] {(C,S,[]:E,s) \Downarrow_{V} (S',l:E) \\ (\GetCtx{C}{(l:[])},S'[(\GetCtx{C}{E},x) \rightarrow \mathit{combine}(l)],E,s') \Downarrow_{V} (S'',E')} {(C,S,E,\CCkeyw{try}\ s\ \CCkeyw{catch}\ x\ \CCkeyw{with}\ s') \Downarrow_{VS} (S'',E')}}
\def \BigVChc {\infer [VS-Chc] {(C \wedge C',S,E,s) \Downarrow_{V} (S',E') \\ (C \wedge \neg C',S',E',s') \Downarrow_{V} (S'',E'')} {(C,S,E,\tagtree{C'}{s}{s'}) \Downarrow_{VS} (S'',E'')}}

\begin{figure}[H]
\begin{syntax}
\text{(Store)}\\
S &::=& S(C,x) & \textit{Lookup x in S with context C} \\
& | & S[(C,x) \rightarrow u] & \textit{Update x in S to u with context C} \\
\text{(Variational integers)}\\
u &::=& n & \textit{Integer literal} \\
& | & \tagtree{C}{u}{u} & \textit{Choice} \\
\text{(Variational booleans)}\\
v &::=& b & \textit{Boolean literal} \\
& | & \tagtree{C}{v}{v} & \textit{Choice} \\
\text{(Error lists)} \\
l &::=& [] & \textit{Nil} \\
& | & (C,u):l & \textit{Cons} \\
\text{(Error context)} \\
E &::=& [] & \textit{Nil} \\
& | & l : E & \textit{Cons} \\ 
\end{syntax}
\end{figure}

\begin{figure}[H]
\ContinuedFloat
\begin{mathpar}
\BigVSat \and
\BigVUnsat \and
\BigVSkip \and
\BigVAssn \and
\BigVSeq \and
\BigVIf \and
\BigVWhile \and
\BigVThrow \and
\BigVTry \and
\BigVCatch \and
\BigVChc
\end{mathpar}
\caption{Big-step semantics of Variational Imp statements}
\label{fig:vimpstmt}
\end{figure}

%\begin{flalign*}
%& \mathit{ctx}(C,[]) & = &\ C & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & &  & & & & & & & & & & & & & & & & & & & & & & & & & & & & & & &  \\
%& \mathit{ctx}(C,([]:E)) & = &\ \mathit{ctx}(C,E) \\
%& \mathit{ctx}(C,((C',n):xs):ys) & = &\ \mathit{ctx}(C \wedge \neg C',xs:ys)
%\end{flalign*}

>>>>

\section{Exceptions}
\label{sec:except}

@@

We also present the big-step operational semantics for VIMP expressions and statements. In the following sections we will discuss specific elements of these definitions.
We begin by defining variational operations, such as configuration and selection as explained in Chapter \ref{ch:bg}. We then examine the need for maintaining a variational context during execution, especially in its relation to the variable store. We then turn to a discussion of
catchable exceptions. Finally, we discuss interacting with control structures.

\begin{figure}[H]
\begin{syntax}
\text{(Store)}\\
S &::=& S(C,x) & \textit{Lookup x in S with context C} \\
\text{(Variational integers)}\\
u &::=& n & \textit{Integer literal} \\
& | & \tagtree{C}{u}{u} & \textit{Choice} \\
\text{(Variational booleans)}\\
v &::=& b & \textit{Boolean literal} \\
& | & \tagtree{C}{v}{v} & \textit{Choice}
\end{syntax}

\begin{mathpar}
\BigVNum \and
\BigVRef \and
\BigVAdd \and
\BigVAChcOne \and
\BigVAChcTwo \and
\BigVAChcThree
\end{mathpar}
\label{fig:vimpexpr}
\caption{Big-step semantics of Variational IMP expressions}
\end{figure}

\begin{figure}[H]
\ContinuedFloat

\begin{mathpar}
\BigVB \and
\BigVNot \and
\BigVAnd \and
\BigVLess \and
\BigVBChcOne \and
\BigVBChcTwo \and
\BigVBChcThree
\end{mathpar}
\caption{Big-step semantics of Variational IMP expressions (continued)}
\end{figure}

\newcommand{\GetCtx}[2]{\mathit{ctx}(#1,#2)}
\def \gc {\GetCtx{C}{E}}
\def \T {\mathit{true}}
\def \F {\mathit{false}}
\def \BigVUnsat {\infer [V-Unsat] {\mathit{unsat}(\gc)} {(C,S,E,s) \Downarrow_{V} (S,E)}}
\def \BigVSat {\infer [V-Sat] {\mathit{sat}(\gc) \\ (C,S,E,s) \Downarrow_{VS} (S',E')} {(C,S,E,s) \Downarrow_{V} (S',E')}}
\def \BigVSkip {\infer[VS-Skip] { } {(C,S,E,\CCkeyw{skip}) \Downarrow_{VS} (S,E)}}
\def \BigVAssn {\infer[VS-Assn] {(\gc,S,a) \Downarrow_{VA} u} {(C,S,E,x \coloneqq a) \Downarrow_{VS} (S[(\gc,x) \rightarrow u],E)}}
\def \BigVSeq {\infer [VS-Seq] {(C,S,E,s) \Downarrow_{V} (S',E') \\ (C',S',E',s') \Downarrow_{V} (S'',E'')} {(C,S,E,s\ ;\ s') \Downarrow_{VS} (S'',E'')}}
\def \BigVIf {\infer [VS-If] {(\gc,S,e) \Downarrow_{VB} v \\ (C \wedge \mathit{whenTrue}(v),S,E,s) \Downarrow_{V} (S',E') \\ (C \wedge \mathit{whenFalse}(v),S',E') \Downarrow_{V} (S'',E'')} {(C,S,E,\CCkeyw{if}\ e\ \CCkeyw{then}\ s\ \CCkeyw{else}\ s') \Downarrow_{VS} (S'',E'')}}
\def \BigVWhile {\infer [VS-While] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ (C \wedge \mathit{whenTrue}(v),S,E,s) \Downarrow_{V} (S',E') \\ (C \wedge \mathit{whenTrue}(v),S',E',\CCkeyw{while}\ e\ \CCkeyw{do}\ s) \Downarrow_{V} (S'',E'')} {(C,S,E,\CCkeyw{while}\ e\ \CCkeyw{do}\ s) \Downarrow_{VS} (S'',E'')}}
\def \BigVThrow {\infer [VS-Throw] {(\GetCtx{C}{l:E},S,a) \Downarrow_{VA} u} {(C,S,l:E,\CCkeyw{throw}\ a) \Downarrow_{VS} (S,((C,u):l):E)}}
\def \BigVTry {\infer [VS-Try] {(C,S,[]:E,s) \Downarrow_{V} (S',[]:E)} {(C,S,E,\CCkeyw{try}\ s\ \CCkeyw{catch}\ x\ \CCkeyw{with}\ s') \Downarrow_{VS} (S',E)}}
\def \BigVCatch {\infer [VS-Catch] {(C,S,[]:E,s) \Downarrow_{V} (S',l:E) \\ (\mathit{catchCtx}(C,l),S'[(\GetCtx{C}{E},x) \rightarrow \mathit{extract}(l)],E,s') \Downarrow_{V} (S'',E')} {(C,S,E,\CCkeyw{try}\ s\ \CCkeyw{catch}\ x\ \CCkeyw{with}\ s') \Downarrow_{VS} (S'',E')}}
\def \BigVChc {\infer [VS-Chc] {(C \wedge C',S,E,s) \Downarrow_{V} (S',E') \\ (C \wedge \neg C',S',E',s') \Downarrow_{V} (S'',E'')} {(C,S,E,\tagtree{C'}{s}{s'}) \Downarrow_{VS} (S'',E'')}}

\begin{figure}[H]
\begin{syntax}
\text{(Store)}\\
S &::=& S(C,x) & \textit{Lookup x in S with context C} \\
& | & S[(C,x) \rightarrow u] & \textit{Update x in S to u with context C} \\
\text{(Variational integers)}\\
u &::=& n & \textit{Integer literal} \\
& | & \tagtree{C}{u}{u} & \textit{Choice} \\
\text{(Variational booleans)}\\
v &::=& b & \textit{Boolean literal} \\
& | & \tagtree{C}{v}{v} & \textit{Choice} \\
\text{(Error lists)} \\
l &::=& [] & \textit{Nil} \\
& | & (C,u):l & \textit{Cons} \\
\text{(Error context)} \\
E &::=& [] & \textit{Nil} \\
& | & l : E & \textit{Cons} \\ 
\end{syntax}
\caption{Big-step semantics of Variational Imp statements}
\label{fig:vimpstmt}
\end{figure}

\begin{figure}[H]
\ContinuedFloat
\begin{mathpar}
\BigVSat \and
\BigVUnsat \and
\BigVSkip \and
\BigVAssn \and
\BigVSeq \and
\BigVIf \and
\BigVWhile \and
\BigVThrow \and
\BigVTry \and
\BigVCatch \and
\BigVChc
\end{mathpar}
\caption{Big-step semantics of Variational Imp statements (continued)}
\end{figure}

@@

When describing the issues involved in the interaction between mutable state and variation in Section \ref{sec:mutst}, we noted the need for (1) maintaining a variational context
during execution and (2) using this variational context when interacting with the variable store. This is a solution first discovered by \cite{varwhile} in their work on a
variational model of execution for the \textsc{While} language.

@@

@@

vupdate :: Cond -> Var -> V v -> VStore v -> VStore v
vupdate c k v vst = case lookup k vst of
    Nothing -> insert k (Chc c (fmap Just v) (One Nothing))
    Just old -> insert k (Chc c (fmap Just v) old)
\end{lstlisting}
\caption{Variational map operations}
\label{fig:varmap}
\end{figure}

In Figure \ref{fig:varmap} we provide a sample implementation of a variational store based on the variational map data structure \cite{varwhile, walkingshaw2014variational}.
Variational maps are an efficient map data structure for storing variational values and are easily built using an existing non-variational map. The basic idea
is that for a non-variational map of with values of type \prog{v}, a variational map uses the same underlying data structure but with values of type \prog{V (Maybe v)}.
The variational values in the map must be of type \prog{Maybe v} in order to account for cases where a value is assigned in one variant but not in others. Our sample
variational store is built on top of Haskell's \prog{Data.Map} library, but in theory any map data structure should suffice.

In VIMP, each evaluation step requires the current variational context. In VIMP's expressions, the variational context is passed to the lookup operation for any
variable reference. Similarly, in VIMP's statements the variable assignment operation takes the current variational context as an argument.

\section{Variational Exceptions}
\label{sec:varexcept}

Catchable exceptions are the other type of side effect that we want VIMP to support. In Section \ref{sec:except} we outlined the primary requirement for exceptions
in a variational setting: when an exception is thrown in a particular variant, all execution in that variant should stop until the error is (optionally) caught, but other variants
should continue their execution unaffected. In order to achieve this, we utilize the existing framework we have built for tracking the current variational context but with some
modifications to account for the presence of exceptions.

If we wish to stop execution in a variant when an exception is thrown, then the $\CCkeyw{throw}$ operation needs to be capable of returning its current variational context
along with the specific error code of the exception. We have already incorporated a variational context into the execution of VIMP, so it is a trivial extension to have $\CCkeyw{throw}$
return this value. The variational
context framework also easily supports preventing the continued execution of variants that are in an error state. We simply need to incorporate an error context into our
satisfiability checks such that the variational context is unsatisfiable for variants in an error state. This means that before beginning execution on a new variant we now check
for (1) whether the given configuration is even possible and (2) whether the variant is currently in error and should not be executed. The error context can be easily derived
by taking the disjunction of the variational contexts returned by all $\CCkeyw{throw}$ operations. To make entering an error context impossible we take the conjunction of
the current variational context and the negation of the error context and check it for satisfiability before entering a new variant.

Supporting catchable exceptions gives us several additional challenges. We need to be able to ``roll back" the error context to a previous state. This suggests that
during execution the error context and corresponding values should be kept in a stack. When we enter a new $\CCkeyw{try}$ block we push an empty error frame onto
the stack. All error values and their contexts thrown in the execution of the $\CCkeyw{try}$ block get added to the top of the stack. We pop the top frame of the stack when
we reach the corresponding $\CCkeyw{catch}$ block. If we reach the $\CCkeyw{catch}$ handler with
a non-empty top stack frame, we then need to (1) extract the variational value of all of the error codes thrown during the try block and assign it to the provided variable name and (2) execute the catch block from within the
combined variational context that those exceptions were thrown in.

\begin{figure}
\begin{lstlisting}
type Err = (Cond, V Int)
type Stack a = [a]

ctx :: Cond -> Stack [Err] -> Cond
ctx c st
    | [] <- concat st = c
    | errs <- concat st = c :&&: f errs
  where
    f [(c,_)] = c
    f ((c,_):es) = c :||: f es

catchCtx :: Cond -> [Err] -> Cond
catchCtx c errs = c :&&: foldl f (Lit False) errs
  where
    f c (c',_) = c :||: c'

extract :: [Err] -> V Int
extract [] = error "no values to extract"
extract [(_,v)] = v
extract ((c,v):es) = Chc c v (extract es)
\end{lstlisting}
\caption{Variational error helper functions}
\label{fig:errhelpers}
\end{figure}

With these requirements in mind, we define the error context carried during execution to be a stack of lists of pairs of variational contexts and error codes.
We also define the helper functions $\mathit{ctx}$, $\mathit{catchCtx}$ and $\mathit{extract}$. $\mathit{ctx}$ retrieves the condition for the satisfiability check
before entering a new variant by combining the error context and the variational context. $\mathit{catchCtx}$ determines the correct variational context for a $\CCkeyw{catch}$
block. $\mathit{extract}$ creates a variational value from all of the error codes thrown in a try block. Haskell implementations of these functions are presented in Figure \ref{fig:errhelpers}.

\section{Control Structures}
\label{sec:control}

Control structures are another aspect of VIMP where the interaction with variation plays a crucial role. IMP's control structures---$\CCkeyw{if \ldots then \ldots else}$ and $\CCkeyw{while \ldots do}$---conditionally execute statements based on the evaluation of a boolean expression. In VIMP, this presents two issues. First, any variation in the condition of a control structure
necessarily affects the conditionally executed block. Second, a formula tree representing a variational boolean
value may contain multiple distinct $\CCkeyw{true}$ or $\CCkeyw{false}$ values. With most other operations in VIMP such as addition or comparison we simply lift them into the applicative functor for formula trees.
However, when interacting with control structures this approach seems sure to cause significant inefficiencies. The applicative functor simply unwraps each leaf value in the tree
and continues execution. This might lead us to reevaluate the same conditional block repeatedly, only with different variational contexts. Any sharing between executions of this
conditional block would be lost.

The first issue is easily resolved by adding the context that caused the boolean value to be $\CCkeyw{true}$ (or in the case of $\CCkeyw{else}$ blocks, $\CCkeyw{false}$)
to the variational context for the conditionally executed statements. For example, the statement
$\CCkeyw{if}\ \tagtree{A}{\CCkeyw{true}}{\CCkeyw{false}}\ \CCkeyw{then}\ s\ \CCkeyw{else}\ s'$, the statement $s$ would be evaluated with variational context $A$, while
$s'$ would be evaluated under $\neg A$.

For the second issue, we need a way to extract a condition that encompasses all variants where a variational boolean value is $\CCkeyw{true}$ or, in the case of $\CCkeyw{else}$ blocks,
$\CCkeyw{false}$. For this, we define two functions, $\mathit{whenTrue}$ and $\mathit{whenFalse}$ that return these conditions for a given variational boolean. We then use these
values
